\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}

### 2021-08-05 move_schedules.tex
Finished the natural images section! Looking much better

If one was not aware of the analysis done by~\cite{lee2003nonlinear, hateren_schaaf_1998, carlsson2008local, perea2014klein}, it is not immediately clear a priori that the Klein bottle model is indeed an excellent candidate for capturing the non-linearity of the data. Nonetheless, as we have mentioned, applications of persistent homology detecting the presence of a topological space persistent homology is designed to reveal. Consider the (coarsened) fibered barcode shown in Figure~\ref{fig:patch_data_dgm} which was computed from a standard Rips/codensity bifiltration on a constructed on a representative sample of the image data from~\cite{hateren_schaaf_1998} (how we obtained this sample is described further down). Upon inspection of the bigraded Betti number and the dimension function, one finds that a large area of dimension function is constant (highlighted as the blue portion in the middle of Figure~\ref{fig:patch_data_dgm}), wherein $\mathrm{dim}(M) = 5$. This indicates

Using the same high-contrast patch data set studied in~\cite{lee2003nonlinear}, we evaluate the performance of various methods at computing the fibered barcode invariant using the parameterization from~\ref{sec:fibered_barcode_reparam}. First, we give a summary of the data set. 
Specifically, we begin by extracting a representative subset of the set $X \subset \mathcal{X}$ of patches representing the top 30\% densest patches of the space of all patches $\mathcal{X}$, where the notion of density is approximated with a k-nearest neighbor (knn) density estimator $\rho$. Carlsson et al.~\cite{carlsson2008local} studied a variety of choices of $k$ in order to understand how localized the data were distributed around the Klein bottle model. He determined empirically that when $k \approx 15$, the top 30\% captures three types of directional shifts in the grayscale gradients of the image patches. These three gradients were shown to be captured well by the three-circle model $C_3$---the 1-skeleton of the Klein bottle. The $C_3$ model consists of three circles, two of which (say, $S_v$ and $S_h$) intersect the third (say, $S_\mathrm{lin}$) in exactly two points, but themselves do not intersect. One of the circles, $S_{\mathrm{lin}}$, parameterizes a the rotational gradient, whereas the other two $S_v$ and $S_h$ parameterize vertical and horizontal gradients.  
A projection of the data $X$ is shown in the left-most image in Figure~\ref{fig:patch_data_dgm}, along with 15 landmark points. It was shown in~\cite{carlsson2008local} that when $k \approx 300$, the top 30\% of data extracted from the knn density estimator $\rho_{300}$ `fills out' the two-manifold, yielding Betti numbers $\beta_0 = \beta_1 = 1$, whereas when $k$ is decreases to a more localized estimate, the three-circle model is recovered, which has Betti numbers $\beta_0 = 1, \beta_1 = 5$.

Due to the aformentioned high complexity of the fibered barcode computation, we begin by working with a subset of the image patch data $\mathcal{X}$. Specifically, we combine the use furthest-point sampling and proportionate allocation (stratified) sampling to sample landmarks $X \subset \mathcal{X}$ distributed within $n = 25$ strata. Each strata consists of the $(1/n)$-thick level set given by $\rho_{15}$, the density estimator. The use of furthest-point sampling gives us certain coverage guarantees that the geometry is approximately preserved within each level set, whereas the stratification ensures the original density of is approximated preserved as well. From this data set, we construct a Rips-(co)density bifiltration $\mathcal{F}$ using $\rho_{15}$ equipped with the geodesic metric computed over the $k = 15$ knn graph on $X$.

In theory, this bifiltration should be well approximated by the $C_3$ model, which one ought to be able to determine by a careful inspection of the fibered barcode. That is, since $C_3$ has $\beta_1  = 5$, there should be some line $L$ which induces a 1-parameter filtration whose corresponding persistence diagram (for dimension $1$) displays five highly persistent features. We show an example of the data, the bigraded Betti numbers, and the persistence diagram extracted from a particular choice of line $L$ in Figure~\ref{fig:patch_data_dgm}.

The example from~\ref{fig:patch_data_dgm} demonstrates a practical application of the rank invariant to uncovering the underlying topological structure characterizing a ``real-world'' data set. Indeed, this is just one of the many applications of multidimensional persistence. As discussed, one of the main issues inhibiting the use of rank invariant in practice is computation. In Table~\ref{tab:barcode_templates}, we record a variety of statistics related to the computation of the rank invariant on the bifiltration constructed from the natural images data set. 
TODO: finish explanation
%An increasingly refined

\begin{table}[h]
\caption{Cost to compute all barcode templates $\mathcal{T}$ for various coarsening choices $\beta(M)$.}\label{tab:barcode_templates}
\centering
\begin{tabular}{ |P{1.6cm}|P{1.6cm}|P{1.2cm}|P{2.8cm}|P{2.4cm}|}
 \hline
 %\multicolumn{7}{|c|}{ Size and cost of constructing $\mathcal{A}^\ast(M)$ } \\
 %\hline
  \multicolumn{1}{|c|}{ $\beta(M)$ } & \multicolumn{1}{|c|}{$\mathcal{A}(M)$ } & \multicolumn{3}{c|}{ Column reductions / Permutations } \\
 \hline
 Coarsening & \# 2-cells & \texttt{phCol} & Vineyards / $d_{K}$ & Moves / $d_{\text{LCS}}$ \\
 \hline
 8 x 8 & 39 & 94.9K & 245K / 1.53M & 38.0K / 11.6K   \\
 \hline
 12 x 12 & 127 & 318K & 439K / 2.66M & 81.9K / 33.0K \\
 \hline 
 16 x 16 & 425 & 1.07M & 825K / 4.75M & 114K / 87.4K \\
 \hline 
  20 x 20 & 926 & 2.32M & 1.15M / 6.77M  & 148K / 154K\\
 \hline 
  24 x 24 & 1.53K & 3.92M & 1.50M / 8.70M & 184K / 232K  \\
   \hline
 \end{tabular}
\label{table:dual_graph_costs}
\end{table}

>>>>

If one was not aware of the analysis done by~\cite{lee2003nonlinear, hateren_schaaf_1998, carlsson2008local, perea2014klein}, it is not immediately clear a priori that the Klein bottle model is indeed a good candidate for capturing non-linearity of image patches. Nonetheless, as we have mentioned, among the first uses of persistent homology was the so-called ``homology inference problem''~\cite{perea2018brief}; given a suitable treatment of strong outliers using by e.g. filtering both on the geometry and the density of the image patches, the homology of the Klein bottle should become apparent through a persistence computation.

Consider the (coarsened) fibered barcode computed from a standard Rips/codensity bifiltration on a representative sample of the image data from~\cite{hateren_schaaf_1998} (details to follow), shown in Figure~\ref{fig:patch_data_dgm}. 
Upon inspection of the bigraded Betti number and the dimension function, one finds that a large area of dimension function is constant (highlighted as the blue portion in the middle of Figure~\ref{fig:patch_data_dgm}), wherein the first Betti number is 5. There are many spaces whose  meets this criterion, though upon further inspection, one candidate space that seems particularly plausible is the three-circle model $C_3$. The $C_3$ model consists of three circles, two of which (say, $S_v$ and $S_h$) intersect the third (say, $S_\mathrm{lin}$) in exactly two points, but themselves do not intersect. Upon performing an analysis on the density of the patches analogous to the one done by~\cite{lee2003nonlinear}, one finds that the mean-centered and contrast normalized $3 \times 3$ patches like on a 7-dimensional ellipsoid $\tilde{S}^7 \subset \mathbb{R}^9$. Contrast is typically measured using a discrete version of the Dirichlet semi-norm, or the ``$D$-norm'' $\lVert \cdot \rVert_D$ for short, which is a unique scale-invariant norm on images:
$$ \lVert x \rVert_D = \sqrt{\sum_{i \sim j}(x_i - x_j)^2} = \sqrt{x^T D x}$$
where $D$ is a fixed matrix which upon application to an image $x \in \mathbb{R}^9$ yields a value proportional to the sum of the differences between the 4-connected neighbors (described by the relation $i \sim j$). It is advantageous to ``whiten'' the data via a change of coordinates to Euclidean sphere. 
This can be achieved by expressing the patch data as an expansion of a certain convenient basis of 8 non-constant eigenvectors, often called the Discrete Cosine Transform (DCT) basis~\cite{lee2003nonlinear}, which are given below: 
\vspace{-0.2em}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.60\textwidth]{dct_basis}
\end{figure}
\vspace{-0.5em}

\noindent Projecting the image data onto the first two basis vectors of leads to the projection shown in the top left of Figure~\ref{fig:patch_data_dgm}. 15 landmark points are also shown. Observe the data are distributed well around three ``circles''---the outside circle capturing the rotation gradient of the image patches ($S_{\mathrm{lin}}$), and the other two circles capturing the vertical and horizontal gradients ($S_v$ and $S_h$, respectively). Since the three circle model is the 1-skeleton of the Klein bottle~\cite{carlsson2008local}, one may conclude that the Klein bottle might be a reasonable candidate upon which the image data are distributed. This can be further verified empirically by e.g. using the iterative, meanshift-like procedure described in~\cite{carlsson2008local} to fit a parameterization of the model in question to the data.

Using the same high-contrast patch data set studied in~\cite{lee2003nonlinear}, we evaluate the performance of various methods at computing the fibered barcode invariant using the parameterization from~\ref{sec:fibered_barcode_reparam}. 
Due to the aformentioned high complexity of the fibered barcode computation, we begin by working with a subset of the image patch data $\mathcal{X}$. Specifically, we combine the use furthest-point sampling and proportionate allocation (stratified) sampling to sample landmarks $X \subset \mathcal{X}$ distributed within $n = 25$ strata. Each strata consists of the $(1/n)$-thick level set given by $\rho_{15}$, the density estimator. The use of furthest-point sampling gives us certain coverage guarantees that the geometry is approximately preserved within each level set, whereas the stratification ensures the original density of is approximated preserved as well. From this data set, we construct a Rips-(co)density bifiltration using $\rho_{15}$ equipped with the geodesic metric computed over the same $k = 15$ knn graph on $X$. 
Finally, we record the number of column reductions needed to compute the fibered barcode at a variety of levels of coarsening using $\mathtt{pHcol}$, vineyards, and our moves approach. The results are summarized in Table~\ref{tab:barcode_templates}. We also record the number of 2-cells encountered along the path $\Gamma$ permutations that were applied throughout the computation of the barcode templates for both vineyards and moves, denoted in the table as $d_K$ and $d_{\mathrm{LCS}}$, respectively. 
%Specifically, we begin by extracting a representative subset of the set $X \subset \mathcal{X}$ of patches representing the top 30\% densest patches of the space of all patches $\mathcal{X}$, where the notion of density is approximated with a k-nearest neighbor (knn) density estimator $\rho$. Carlsson et al.~\cite{carlsson2008local} studied a variety of choices of $k$ in order to understand how localized the data were distributed around the Klein bottle model. He determined empirically that when $k \approx 15$, the top 30\% captures three types of directional shifts in the grayscale gradients of the image patches. These three gradients were shown to be captured well by the three-circle model $C_3$---the 1-skeleton of the Klein bottle.  One of the circles, $S_{\mathrm{lin}}$, parameterizes a the rotational gradient, whereas the other two $S_v$ and $S_h$ parameterize vertical and horizontal gradients.  
%It was shown in~\cite{carlsson2008local} that when $k \approx 300$, the top 30\% of data extracted from the knn density estimator $\rho_{300}$ `fills out' the two-manifold, yielding Betti numbers $\beta_0 = \beta_1 = 1$, whereas when $k$ is decreases to a more localized estimate, the three-circle model is recovered, which has Betti numbers $\beta_0 = 1, \beta_1 = 5$.
%In theory, this bifiltration should be well approximated by the $C_3$ model, which one ought to be able to determine by a careful inspection of the fibered barcode. That is, since $C_3$ has $\beta_1  = 5$, there should be some line $L$ which induces a 1-parameter filtration whose corresponding persistence diagram (for dimension $1$) displays five highly persistent features. We show an example of the data, the bigraded Betti numbers, and the persistence diagram extracted from a particular choice of line $L$ in Figure~\ref{fig:patch_data_dgm}. 
%The example from~\ref{fig:patch_data_dgm} demonstrates a practical application of the fiber barcode  to uncovering the underlying topological structure characterizing a ``real-world'' data set. Indeed, this is just one of the many applications of multidimensional persistence. As discussed, one of the main issues inhibiting the use of rank invariant in practice is computation. In Table~\ref{tab:barcode_templates}, we record a variety of statistics related to the computation of the rank invariant on the bifiltration constructed from the natural images data set. 
%TODO: finish explanation
%An increasingly refined 
\begin{table}[h]
\caption{Cost to computing $\mathcal{T}$ for various coarsening choices of $\beta(M)$.}\label{tab:barcode_templates}
\begin{tabular}{ |P{1.6cm}|P{1.6cm}|P{1.2cm}|P{2.8cm}|P{2.4cm}|}
 \hline
 %\multicolumn{7}{|c|}{ Size and cost of constructing $\mathcal{A}^\ast(M)$ } \\
 %\hline
  \multicolumn{1}{|c|}{ $\beta(M)$ } & \multicolumn{1}{|c|}{$\mathcal{A}(M)$ } & \multicolumn{3}{c|}{ Column reductions / Permutations } \\
 \hline
 Coarsening & \# 2-cells & \texttt{phCol} & Vineyards / $d_{K}$ & Moves / $d_{\text{LCS}}$ \\
 \hline
 8 x 8 & 39 & 94.9K & 245K / 1.53M & 38.0K / 11.6K   \\
 \hline
 12 x 12 & 127 & 318K & 439K / 2.66M & 81.9K / 33.0K \\
 \hline 
 16 x 16 & 425 & 1.07M & 825K / 4.75M & 114K / 87.4K \\
 \hline 
  20 x 20 & 926 & 2.32M & 1.15M / 6.77M  & 148K / 154K\\
 \hline 
  24 x 24 & 1.53K & 3.92M & 1.50M / 8.70M & 184K / 232K  \\
   \hline
 \end{tabular}
\label{table:dual_graph_costs}
\end{table}

\noindent
As shown on the table, we're able to achieve a significant reduction in the number of total column operations needed to compute $\mathcal{T}$ compared to both vineyards and $\mathtt{pHcol}$, though it's worth noting that permutations are constant time for vineyards while they require at most $O(m)$ elementary operations for moves. Observe that initially, when there is a high degree of coarsening, vineyards is particularly inefficient and moves requires only about 3x less column operations that naively computing $\mathcal{T}$ independently. As the coarsening becomes more refined and more 2-cells are added to $\mathcal{A}(M)$, however, vineyards quickly becomes a much more viable option compared to $\mathtt{pHcol}$, though even at the highest coarsening we tested the gain in efficient was relatively small. In contrast, our proposed moves approach scales quite well with this refinement, requiring about $12\%$  and $\approx 5\%$ of the number of column operations as vineyards and $\mathrm{pHcol}$, respectively.
